---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    toc: yes
    fig_caption: true
    latex_engine: pdflatex
    template: textemplate_svmiller.tex
title: "A Pandoc Markdown Article Template modified for R-Ladies Berlin"
thanks: "Replication files are available on the author's Github account (http://github.com/svmiller). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: svmille@clemson.edu."
author:
- name: Steven V. Miller
  affiliation: Clemson University
- name: Christin
  affiliation: R-Ladies Berlin, Germany
abstract: "This document provides an introduction to R Markdown
and presents a sample manuscript template intended for an academic audience. 
R-Ladies Berlin added some example plots to illustrate the power of R markdown and pandoc."
keywords: "pandoc, r markdown, knitr, R-Ladies Berlin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
#spacing: double
---

# Introduction

Pandoc is a universial document converter.

Pandoc understands a number of useful markdown syntax extensions, including document metadata 
(title, author, date); footnotes; tables; definition lists; superscript and subscript; strikeout; 
enhanced ordered lists (start number and numbering style are significant); running example lists; 
delimited code blocks with syntax highlighting; smart quotes, dashes, and ellipses; markdown 
inside HTML blocks; and inline LaTeX. If strict markdown compatibility is desired, all of these 
extensions can be turned off.

# Getting started

## Installation

An extensive introduction to command line pandoc and installation recommendations are shown at 
[Pandoc page](http://pandoc.org/getting-started.html).


In generell it requires the installation of a latex distribution for rendering pdf documents,
e.g., [Miktex](https://miktex.org).

------

## YAML extensions

Optional extensions are listed in R markdown cheat sheet, e.g, 

- author affiliations
- content tables
- bibliography
- abstract

## An R code example

Let's include an example plot of the diamonds dataset and play with figure caption,
size and table output formats.


The theme has been developed by Samantha Tyner for R-Ladies global. It modifies
the general ggplot2 settings.

The following modifications are included:
  *  text color changed to purple
  *  bold axis description
  *  grid color: grey 
  *  legend title: bold and purple
  *  plot title: purple, bold, slightly bigger x1.4
  
  
```{r rladies_theme, tidy=TRUE}
library(ggplot2)

#make plots nicer
# R-Ladies ggplot2 theme from Samantha Tyner

r_ladies_theme <- function(base_size){
  theme_bw(base_size) %+replace%
    theme(axis.text = element_text(colour = "#181818"),
          axis.title = element_text(face = "bold", 
              colour = "#88398A", size = rel(1.1)),
          plot.title = element_text(face = "bold", size = rel(1.4),
                                    colour = "#181818",
                                    margin = margin(t = 0, r = 0, b = 6.6,
                                                    l = 0, unit = "pt")),
          legend.title = element_text(face = "bold", colour = "#181818"),
          panel.grid.major = element_line(color = "#D3D3D3"))
}

theme_set(r_ladies_theme(base_size = 10))

```

Let's create a simple plot based on the diamonds dataset. If you want to know more 
about the dataset have a look ??diamonds.

```{r example_plot, fig.cap='Example plot of diamonds data illustrating the price of diamonds per cut.'}

ggplot(diamonds, aes(cut, price, color = cut))+
  geom_jitter()+
  ylab('Price in ($)')+
  xlab('Cut class')

```

Here you can see how to print beautiful tables. Check out the different formats:
markdown, html, latex, pandoc or rst.

```{r example-table01, results='asis', tidy=TRUE, echo=FALSE}

knitr::kable(head(diamonds), format = "latex", 
  caption = "Just the first rows of the dataset diamonds")

```

