---
title: "playground"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# RethinkR

Lets see what we come up with 

## Scenario 1

### 1. Collective Coding
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```

...


### 2. Loading Packages
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```


### 3. Set missings to NA
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```


### 4. Summary Statistics
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```

### 5. Deflated income
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```


### 6. For loops
```{r}

```

#### Suggest 1
```{r}

```

#### Suggest 2
```{r}

```






## Scenario 2
```{r}

```

#### Suggest 1
bar chart by drink
```{r}
ggplot(data = alcohol_long,
       mapping = aes(x = characteristic, y = approval, fill = drink)) +
  geom_bar(stat = "identity", position = "dodge") +
  #facet_grid(~ drink) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  coord_flip()
```

#### suggest2
```{r}
good <- c("selfconfident", "relaxed", "energetic", "attractive (sexy)")
bad <- c("tired", "aggressive", "bad", "restless", "sad")

# Dataset for plot
p.alc <- alcohol_long %>% 
  mutate(good = ifelse(X.1 %in% good, 1,0),
         bad = ifelse(X.1 %in% bad,1,0),
         approval2 = ifelse(good == 0, approval*(-1), approval)
  ) 

# For each kind of alcohol, how positive or negative are the effects
ggplot(data = p.alc, mapping = aes(x = characteristic, y = approval2, fill = ..y.. > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~ drink) +
  scale_fill_manual(name = "Effect of Alcohol",
                    values = c('red', 'green'),
                    labels = c("negative", "positive"))+
  theme(axis.text.x = element_text(angle = 20, hjust = 1))


# For each effect, how positive or negative is each alcohol
ggplot(data = p.alc, mapping = aes(x = drink, y = approval2, fill = ..y.. > 0)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(~ characteristic) +
  scale_fill_manual(name = "Effect of Alcohol",
                    values = c('red', 'green'),
                    labels = c("negative", "positive"))+
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

#### suggest 3
 ideas:
 - how much approval for positive modds as lineplot?
 - how much approval for negative modds as lineplot
