---
title: "dplyr excercises 1"
output:
  #html_document: default
  pdf_document: default
  #output: html_notebook
urlcolor: blue
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse)
#we use the diamonds-data 
```

##1. Question

Select all Diamnons with the color 'H' and calculate the the number of diamonds, depending on the clarity


```{r, include=FALSE}

diamonds %>% filter(color=='H') %>%  group_by(clarity) %>% summarise(num_diamonds=n())

```



##2. Question

Select all diamonds with a number of carats bigger than 3. Calculate the mean, median and sum of price and the number of stones depending on the cut. Order the data depending on the number of stones.

```{r, include=FALSE}

diamonds %>% filter(carat>3) %>% group_by(cut) %>% summarise(mean_price = mean(price),
                                                             median_price = median(price),
                                                             sum_price = sum(price),
                                                             n=n()) %>% 
                                                    arrange(n)

```

##3. Question

select from all diamonds with carat bigger than 3.5 the columns carat, color and price. Arrange them by price. 
Add two new columns to the data set:
  - the cumulated sum of the price and
  - how much differes the price of a stone from the mean (in percent)


```{r, include=FALSE}

diamonds %>% select(carat, color, price) %>% 
            filter(carat>3.5) %>% arrange(price) %>% 
            mutate(cumsum = cumsum(price),
                  diff_mean = ((price/mean(price))-1)*100)
```
