---
title: "dplyr tutorial"
output:
  #html_document: default
  #pdf_document: default
  output: html_notebook
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dplyr is a grammar for data manipulation. It combines various steps of manipulation in one chunk of code without nesting. 

For more information check the official [webside](http://dplyr.tidyverse.org/).

##Tidyverse
The package is part of the "tidyvers-Universe". 
Tidyverse is a a collection of packages which share a common philosophy of data analysis and R programming and are designed to work together seamlessly.

- ggplot2, for data visualisation.
- dplyr, for data manipulation.
- tidyr, for data tidying.
- readr, for data import.
- purrr, for functional programming.
- tibble, for tibbles, a modern re-imagining of data frames.

They share common data representation and API design and the tidyverse package allows you to install an update them as a set (they update frequently).


##Introduction
This tutorial will introduce the basic commands to work with dplyr.
This [cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf)by R-Studio gives a overview of more helpful commands.

The functions work in a "regular" way and in the %>%-version. 

  + function(data,....)

or

  + data %>% function(....).

As the author of this tutorial likes the %>% a lot, most functions will be shown in this way.
The shortcut for ' %>% ' is strg+command+m  
<br>
<br>
Let's get started. First you have to install and load the the tidyverse package (or dplyr and ggplot2 if you want to do it separatly). 

```{r message=FALSE}
library(tidyverse)
#we use the diamonds-data 
```


##select()
selects a set of columns

```{r}

#Classic Style
select(diamonds, cut, color) 

```
```{r}
#pipe-style
diamonds %>% select(cut,color) 
```

```{r}
#adding the comand head at the end, to only select the fist five rows
diamonds %>% select(cut, color) %>% head
```

```{r}
#select all rows staring with "c" 
diamonds %>% select(starts_with('c'),price)

#Also works with ends_with(), contains(), matches(),...
```

#filter
You can filter by the values of one or more columns.

```{r}
#filter by one column
diamonds %>% filter(color=='E')
diamonds %>% filter(price>350)
```

```{r}
#the %in%-command is usefull to filter on several values
diamonds %>% filter(color %in% c('E','F'))
```

```{r}
#The ! in front of the command changes the operator to "not in"
diamonds %>% filter(!color %in% c('E','F')) 
```

```{r}
#filter on more than one column

#The "AND"-combination can be done in two ways
diamonds %>% filter(color=='E' , price>350) 

#diamonds %>% filter(color=='E' & price>350)
#has the same output

```

```{r}
#filtering with "OR" on two columns
diamonds %>% filter(color=='E' | price>350)


```

#arrange 
reorder the rows by one or more columns
```{r}
#reorder by the column "cut"
diamonds %>% arrange(cut)
```

```{r}

#reorder by the column cut and price. To order in descending way, put the column in desc()
diamonds %>% arrange(cut,desc(price)) 


```
#mutate 
you can create new columns via mutate

```{r}
#simple new column
diamonds %>% select(cut, color, price) %>% 
              mutate(price_n = price*1000)
```

```{r}

#new column base on a case-statement
diamonds %>% select(cut, color, price) %>% 
              mutate(cut_n = case_when(.$cut=='Ideal' ~ '1',
                                      .$cut=='Premium' ~ '2',
                                      .$cut=='Very Good' ~ '3',
                                      .$cut=='Good' ~ '4',
                                      .$cut=='Fair' ~ '5',
                                      TRUE ~ '9'))
#if you miss a value, it is going to be NA
# the TRUE-statement ist kind of like a "else"-statement

```
#summarise 

You can use all kind of functions to summarize the data (sum, mean, median, .....). 

```{r}
#calculating the mean of the column price
diamonds %>% summarise(mean(price))
```

```{r}

#adding the number distinct values of the column cut and the number of rows in the dataframe
diamonds %>% summarise(mean_price=mean(price),
                       num_cut=n_distinct(cut),
                       num_row=n())

```

#group_by 
group_by builds groups by the defined columns and does the operations within those groups.
For those familiar with sql it works pretty much the same as "group by"

```{r}
#Group by one column and then summarising with those groups
diamonds %>% group_by(cut) %>% summarise(mean_price=mean(price))
```

#piping comands together
So fare the commands mostly just looked different to "regular" r-commands. Maybe are more straight forward. But to experience (IMO) the real advantage of the dplyr-package, you have to start piping the commands. 

```{r}

# Select all diamonds with a fair cut and then calucate the mean price depending on the clarity
diamonds %>% filter(cut=='Fair') %>% group_by(clarity) %>% summarise(mean_price=mean(price))


```


# some more helpfull functions

##cast / melt -> functions of the reshaping2 package
If you don't know the functions, check them out. They really help to reorganize data.
<br>
in this context (tidyvers-universe) they are called spread and gather and are part of the package "tidyr"
```{r}

#spread the result 

diamonds %>% 
  filter(cut!='Fair') %>% 
  group_by(cut,clarity) %>% 
  summarise(mean_price=mean(price)) %>% 
  spread(key=cut,value=mean_price)
```

```{r}


#reversing the spread with gather
diamonds %>% 
  filter(cut!='Fair') %>% 
  group_by(cut,clarity) %>% 
  summarise(mean_price=mean(price)) %>% 
  spread(key=cut,value=mean_price)  %>% 
  gather(cut,price, Good:Ideal )
  
```


##show the entire data.frame
dplyr always shows the first 10 rows, but sometimes you just want to see them all

```{r}
#to see the result in the console, use print(n=nrow(.))
diamonds %>% 
  filter(cut!='Fair') %>% 
  group_by(cut,clarity) %>% 
  summarise(mean_price=mean(price)) %>% 
  spread(key=cut,value=mean_price)  %>% 
  gather(cut,price, Good:Ideal ) %>% 
  print(n = nrow(.))
```

```{r}

#if you use View() RStudio opens a new tab with the data
diamonds %>% 
  filter(cut!='Fair') %>% 
  group_by(cut,clarity) %>% 
  summarise(mean_price=mean(price)) %>% 
  spread(key=cut,value=mean_price)  %>% 
  gather(cut,price, Good:Ideal ) %>% 
  View()


```


##some helpfull other tutorials

- [Data Manipulation with dplyr](https://www.r-bloggers.com/data-manipulation-with-dplyr/)
- [Lesser known dplyr tricks](https://www.r-bloggers.com/lesser-known-dplyr-tricks/)
- [Some tricks on dplyr::filter](https://sebastiansauer.github.io/dplyr_filter/)



Have fun exploring your data!





