---
title: 'Solutions - #1 Hands-on data vis w/ ggplot2'
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

# How many ways exist to the same plot?

```{r init_libraries, include=FALSE}
library(ggplot2)

#another useful packages
library(RColorBrewer)


#set path for figure export and size
knitr::opts_chunk$set(fig.width=7, fig.align='center', fig.height=5, fig.path='HandsOn_ggplot2-',  
                      echo=TRUE,  #show R code
                      warning=FALSE, #show warnings
                      message=TRUE) #show messages


#make plots nicer
# R-Ladies ggplot2 theme from Samantha Tyner

r_ladies_theme <- function(base_size){
  theme_bw(base_size) %+replace% 
     theme(#text = element_text(family = "HelveticaNeue", face = "plain", size = base_size,
    #                           colour = 'black',
    #                           hjust = .5, vjust = .5, angle = 0, 
    #                           lineheight = 1.1, 
    #                           margin = margin(t = 0, r = 0, b = 0, l = 0, 
    #                                           unit = "pt"), 
    #                           debug= FALSE), 
          axis.text = element_text(colour = "#181818"), 
          axis.title = element_text(face = "bold", colour = "#88398A", size = rel(1.1)), 
          plot.title = element_text(face = "bold", size = rel(1.4), 
                                    colour = "#88398A", 
                                    margin = margin(t = 0, r = 0, b = 6.6,
                                                    l = 0, unit = "pt")),
          legend.title = element_text(face = "bold", colour = "#181818"),
          panel.grid.major = element_line(color = "#D3D3D3"))
}

theme_set(r_ladies_theme(base_size = 12))

```


## Plot 1

Description of the plot:

	*x-axis: clarity
	*y-axis: counts
	*grouping: cut (labels right side of graphs)
	*color: cut
	*additional: plots below each other in one column, splitted by cut
	

```{r task_01}

#solution 1
ggplot(diamonds, aes(clarity)) + 
  geom_bar(aes(fill = cut)) +
  facet_grid(cut ~ .)  

#solution 2
ggplot(diamonds, aes(clarity, fill = cut))+
	geom_bar(aes(fill = cut)) +
  facet_grid(cut ~ .)  

#solution 3
ggplot(diamonds, aes(clarity))+
	geom_histogram(stat = 'count', aes(fill = cut)) +
  facet_grid(cut ~ .)  

```


## Task #2

Description of the plot:

	*x-axis: clarity
	*y-axis: counts
	*grouping: cut (labels right side of graphs)
	*color: cut
	*additional: bars next to each other per condition

```{r task_02}

#solution 1
ggplot(diamonds, aes(clarity)) + 
  geom_bar(aes(fill = cut), position = "dodge")

#solution 2
ggplot(diamonds, aes(clarity, fill = cut)) + 
  geom_bar(position = "dodge")

#solution 3
ggplot(diamonds, aes(clarity))+
	geom_histogram(stat = 'count', aes(fill = cut), position = 'dodge') 
```


## Task #3

Description of the plot:

	*x-axis: price
	*y-axis: counts, normalized to 1, stacked
	*grouping: cut
	*binwidth: 500

```{r task_03}

#solution 1
ggplot(diamonds, aes(x=price, fill=cut)) +
	geom_histogram(position="fill", binwidth = 500)

#solution 2
ggplot(diamonds, aes(x=price)) +
	geom_histogram(position="fill", binwidth = 500, aes(fill = cut))

```
